---
title: "Disease effect time uncertainty modeling in lgpr"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Disease effect time uncertainty modeling in lgpr}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

```{r load}
require("lgpr")
require("ggplot2")
```

## Simulating data

In this tutorial we simulate and analyse a test data set which contains 6 case and 6 control individuals, and the disease effect on case individuals is modeled using the disease-related age (`diseaseAge`) as a covariate. The disease-related age is defined as age relative to the observed disease initiation. The true disease effect times for each case individual $q=1, \ldots,6$ are drawn from $\mathcal{N}(36,4^2)$, but the disease initiation is observable only after time $t_q$ , which is drawn from  $t_q∼\text{Exponential}(0.05)$ .

```{r simulate, fig.width = 7, fig.height=5}
set.seed(121)
relev           <- c(0,1,1,1,0,0)
effect_time_fun <- function(){rnorm(n = 1, mean = 36, sd = 4)}
obs_fun         <- function(t){min(t + stats::rexp(n = 1, rate = 0.05), 96 - 1e-5)}
  
simData <- simulate_data(N            = 12,
                         t_data       = seq(12, 96, by = 12),
                         covariates   = c(    0,2,2,2),
                         relevances   = relev,
                         lengthscales = c(18,24, 1.1, 18,18,18),
                         t_effect_range = effect_time_fun,
                         t_observed   = obs_fun,
                         snr          = 3)

plot_sim(simData) + xlab('Age (months)')
#plot_sim(simData, comp_idx = 3) # to visualize one generated component
```
Above, the blue line represents the data-generating signal and black dots are
noisy observations of the response variable.

```{r head}
dat <- simData@data
str(dat)
simData@effect_times
```

## Declaring effect time uncertainty

We will define a formula where the term `unc(id)*gp_vm(diseaseAge)`
declares that the effect time for the nonstationary `gp_vm` term is uncertain and that one uncertainty parameter is needed for each level of `id`.

```{r formula}
formula <- y ~ zs(id)*gp(age) + gp(age) + unc(id)*gp_vm(diseaseAge) + zs(z1)*gp(age) + zs(z2)*gp(age) + zs(z3)*gp(age)
```

Because `diseaseAge` is `NaN`
for the control individuals, it is automatically taken into account that
a separate uncertainty parameter is actually needed just for each case individual.

## Defining effect time prior

We must define a prior for the effect time of each case individual. This means that the `prior` argument must be a list containing elements named `effect_time_info` and `effect_time`. The first one, `effect_time_info`, must be a list, like below:

```{r prior1}
lb <- 0
ub <- c(48,72,96,36,48,60)
effect_time_info <- list(zero = ub, backwards = TRUE, lower = lb, upper = ub)
```

The `lower` and `upper` variables define bounds for the effect times. Here
we have defined the upper bound for each case individual to be the observed disease initiation time, because we know that the true effect has occurred before that. Defining `zero = ub` means that our uncertainty priors are actually for time differences relative to the observed disease initiation time, and `backwards = TRUE` argument is used to define the direction so that the prior is "backwards" in time.

```{r prior2}
my_prior <- list(
  prior = list(wrp = igam(14,5)),
  effect_time_info = effect_time_info,
  effect_time = gam(shape = 1, inv_scale = 0.05)
)
```

The `effect_time` argument now defines the prior distribution for `- (teff - zero)`. If we had `backwards = FALSE` then the specified prior would be for
`teff - zero`.

## Fitting the model

```{r fit, fig.width=7.2, fig.height=4.8, cache = TRUE}
fit <- lgp(formula   = formula,
            data     = dat,
            prior    = my_prior,
            iter     = 2000,
            chains   = 4,
            cores    = 4,
            verbose  = TRUE)
```

Printing the fit object summarizes the posterior
```{r post, fig.width=7.2, fig.height=4.8}
print(fit)
```

Printing the model information clarifies the model and priors
```{r mod}
model_summary(fit)
#rstan::get_elapsed_time(fit@stan_fit)
```

## Visualizing the inferred effect times
We can visualize the inferred effect times for each case individual. We see
that for individuals 2 and 3 the inferred effect time is much earlier than the observed one.
```{r plt1, fig.width=7.2, fig.height=5.2}
plot_effect_times(fit) + xlab('Age (months)')
```

Finally we plot the inferred disease component
```{r plt2, fig.width=7.2, fig.height=4.2}
t <- seq(0, 100, by = 1)
x_pred <- new_x(dat, t, x_ns = 'diseaseAge')
p <- pred(fit, x_pred, verbose = FALSE)
plot_f(fit, x = x_pred, pred = p, comp_idx = 3, color_by = 'diseaseAge')  + xlab('Age (months)')
```

## Computing environment

```{r sess}
sessionInfo()
```
