---
title: "Disease effect time uncertainty modeling in lgpr"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: yeti
vignette: >
  %\VignetteIndexEntry{Disease effect time uncertainty modeling in lgpr}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

```{r load}
require("lgpr")
require("ggplot2")
require("rstan")
```

# Introduction

In this tutorial we simulate and analyse a test data set which contains 6 case
and 6 control individuals, and the disease effect on case individuals is
modeled using the disease-related age (`diseaseAge`) as a covariate. The
disease-related age is defined as age relative to the observed disease
initiation. The true disease effect times for each case individual
$q=1, \ldots,6$ are drawn from $\mathcal{N}(36,4^2)$, but the disease
initiation is observable only after time $t_q$ , which is drawn from
$t_q∼\text{Exponential}(0.05)$ .

```{r simulate, fig.width = 7, fig.height=5}
set.seed(121)
relev <- c(0, 1, 1, 1, 0, 0)
effect_time_fun <- function() {
  rnorm(n = 1, mean = 36, sd = 4)
}
obs_fun <- function(t) {
  min(t + stats::rexp(n = 1, rate = 0.05), 96 - 1e-5)
}

simData <- simulate_data(
  N = 12,
  t_data = seq(12, 96, by = 12),
  covariates = c(0, 2, 2, 2),
  relevances = relev,
  lengthscales = c(18, 24, 1.1, 18, 18, 18),
  t_effect_range = effect_time_fun,
  t_observed = obs_fun,
  snr = 3
)

plot_sim(simData) + xlab("Age (months)")
# plot_sim(simData, comp_idx = 3) # to visualize one generated component
```
Above, the blue line represents the data-generating signal and black dots are
noisy observations of the response variable.

```{r head}
dat <- simData@data
str(dat)
simData@effect_times
```

# Declaring effect time uncertainty

We will define a formula where the term `unc(id)*gp_vm(diseaseAge)`
declares that the effect time for the nonstationary `gp_vm` term is uncertain
and that one uncertainty parameter is needed for each level of `id`.

```{r formula}
formula <- y ~ zs(id) * gp(age) + gp(age) + unc(id) * gp_vm(diseaseAge) + zs(z1) * gp(age) + zs(z2) * gp(age) + zs(z3) * gp(age)
```

Because `diseaseAge` is `NaN` for the control individuals, it is automatically
taken into account that a separate uncertainty parameter is actually needed
just for each case individual.

# Defining the effect time prior

Declaring a temporally uncrertain component will add parameters `teff` to the
model. The vector `teff` has length equal to the number of case individuals.
We must define a prior for each `teff` parameter. This means
that the `prior` argument must be a list containing elements named
`effect_time` and `effect_time_info`. The first one is specified using any
of the basic prior definition functions, like `uniform()`, `normal()`, etc.
The second one, `effect_time_info`, must be a named list containing the fields

 * `zero` - this is a vector with same length as `teff`, and can be used to
 move the center of the prior
 * `backwards` - this is a boolean value, and the prior defined in `effect_time`
 will be for
   * `(teff - zero)` if `backwards = FALSE`
   * `- (teff - zero)` if `backwards = TRUE`
 * `lower` - this is a vector with same length as `teff`, and defines the lower
 bound for each `teff` parameter
  * `upper` - this is a vector with same length as `teff`, and defines the upper
 bound for each `teff` parameter

You can give `zero`, `lower`, and `upper` also as just one number, in which
case they are turned into vectors that repeat the save value. The prior
defined in `effect_time` will be truncated at lower and upper bounds.

## Prior for the effect time directly

We had observed the disease onset at times $48,72,96,36,48,60$ months for each
case individual, respectively. Now if think that the true effect of the
disease has occurred for each indiviaul at some time point before the detection
of the disease, but not before age $18$ months, we could set the prior like
here.

```{r prior0}
obs_onset <- c(48, 72, 96, 36, 48, 60)
lb <- 18
ub <- obs_onset
effect_time_info <- list(zero = 0, backwards = FALSE, lower = lb, upper = ub)
my_prior <- list(
  effect_time = uniform(), # between lb and ub
  effect_time_info = effect_time_info,
  wrp = igam(14, 5) # see how to set this in the 'Basic usage' tutorial
)
```

## Prior relative to a known time point

It is possible that we want a prior where values closer to the observed onset
are more likely than those closer to birth. This can be done by defining
for example an exponentially decaying prior for `-  (teff - obs_onset)`,
as is done here.

```{r prior1}
lb <- 18
ub <- obs_onset
effect_time_info <- list(zero = ub, backwards = TRUE, lower = lb, upper = ub)
my_prior <- list(
  wrp = igam(14, 5),
  effect_time_info = effect_time_info,
  effect_time = gam(shape = 1, inv_scale = 0.05) # = Exponential(rate=0.05)
)
```

Now our uncertainty priors are actually for time differences relative to the
observed disease initiation time, and `backwards = TRUE` argument is used to
define the direction so that the prior is "backwards" in time. We used `gam()`
because the Gamma distribution with `shape=1` and `inv_scale=lambda` is
equal to the Exponential distribution with `rate= lambda`.

# Fitting the model

```{r fit, fig.width=7.2, fig.height=4.8, cache = TRUE}
fit <- lgp(
  formula = formula,
  data = dat,
  prior = my_prior,
  iter = 3000,
  chains = 4,
  cores = 4,
  verbose = TRUE
)
```

Printing the fit object summarizes the posterior
```{r post, fig.width=7.2, fig.height=4.8}
print(fit)
```

Printing the model information clarifies the model and priors
```{r mod}
model_summary(fit)
rstan::get_elapsed_time(fit@stan_fit)
```

# Visualizing the inferred effect times
We can visualize the inferred effect times for each case individual. We see
that for individuals 2 and 3 the inferred effect time is much earlier than the observed one.
```{r plt1, fig.width=7.2, fig.height=5.2}
plot_effect_times(fit) + xlab("Age (months)")
```

Finally we plot the inferred disease component
```{r plt2, fig.width=7.2, fig.height=4.2}
t <- seq(0, 100, by = 1)
x_pred <- new_x(dat, t, x_ns = "diseaseAge")
p <- pred(fit, x_pred, verbose = FALSE)
plot_f(fit, pred = p, comp_idx = 3, color_by = "diseaseAge") + xlab("Age (months)")
```

# Computing environment

```{r sess}
sessionInfo()
```
