---
title: "Using a discrete observation model in lgpr"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Using a discrete observation model in lgpr}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

```{r load}
require("lgpr")
require("ggplot2")
```

## Simulating data

In this tutorial we simulate and analyse count data.

### A) Binomial data 
We first show how to generate binomial data
```{r sim1, fig.width = 7, fig.height=5}
set.seed(131)
simData   <- simulate_data(N            = 12,
                           t_data       = seq(12, 72, by = 12),
                           covariates   = c(    2),
                           noise_type   = "binomial",
                           relevances   = c(0,1,1),
                           lengthscales = c(18,24,18),
                           N_trials     = 100,
                           t_jitter     = 1)

plot_sim(simData)
```


### B) Beta-binomial data 
The noise level in the previous example was rather low. To generate more noisy observations, we can draw from the beta-binomial distribution.
The overdispersion parameter $\gamma \in (0,1)$ controls the noise level,
so that $\gamma \rightarrow 0$ would be equivalent to binomial noise (no overdispersion) and larger values mean more noise. Here we set `gamma = 0.3`.

```{r sim2, fig.width = 7, fig.height=5}
set.seed(131)
simData   <- simulate_data(N            = 12,
                           t_data       = seq(12, 72, by = 12),
                           covariates   = c(    2),
                           noise_type   = "bb",
                           gamma        = 0.3,
                           relevances   = c(0,1,1),
                           lengthscales = c(18,24,18),
                           N_trials     = 100,
                           t_jitter     = 1)

plot_sim(simData)
```

```{r head}
dat <- simData@data
str(dat)
```

## Defining model
To use beta-binomial observation model in our analysis, we use the
`likelihood` argument of `lgp()`. Here we define a $\text{Beta}(2,2)$ prior for
the `gamma` parameter.

```{r prior}
my_prior <- list(gamma = bet(2, 2))
```

## Fitting the model

We fit a model using the generated beta-binomial data.
```{r fit, cache = TRUE}
fit <- lgp(y ~ age + age|id + age|z,
           data = dat,
           likelihood = "bb",
           num_trials = 100,
           prior = my_prior,
           refresh = 300,
           chains = 3,
           cores = 3,
           control = list(adapt_delta = 0.95),
           iter = 2000)
```

```{r post1, cache = TRUE}
print(fit)
```

We see that the posterior mean of the `gamma` parameter is close to the value used in simulation (0.3). We visualize its posterior draws for each MCMC chain:
```{r post2, cache = TRUE, fig.width = 7, fig.height=3.5}
plot_draws(fit, type = "trace", regex_pars = "gamma")
```

## Visualising predictions and components

We subsample 50 draws from the posterior
```{r random}
draws <- sample.int(1000, 50)
```

and use them to visualize the model predictions
```{r pred}
t <- seq(2, 100, by = 2)
x_pred <- new_x(dat, t)
p <- pred(fit, x_pred, draws = draws)
```

```{r plt1, fig.width = 7, fig.height=4.5}
plot_pred(fit, x = x_pred, pred = p, alpha = 0.2)
```

and model components
```{r plt2, fig.width = 7, fig.height=4.5}
plot_components(fit, x = x_pred, pred = p, alpha = 0.1, color_by = c(NA, NA, "z", "z"))
```


## Computing environment

```{r sess}
sessionInfo()
```