---
title: "Basic usage of the lgpr package"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Basic usage of the lgpr package}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

This vignette demonstrates how to model a longitudinal data set and perform covariate relevance assessment using the `lgpr` package.

```{r load}
require("lgpr")
require("ggplot2")
```

```{r simulate, include = FALSE}
set.seed(932)
sim <- simulate_data(N            = 12,                     
                     t_data       = seq(12, 96, by = 12),
                     covariates   = c(    0,2), # covariate types (2 = cat)
                     lengthscales = c(16,24,1,16), # true effect lengthscales
                     relevances   = c(0,1,1,1), # true covariate relevances
                     names        = c("diseaseAge", "sex"),
                     t_jitter     = 0,  # jitter in time points
                     snr          = 3)    # signal-to-noise ratio
dat <- sim@data
dat$y <- 100*(dat$y + 5)
a <- as.numeric(dat$id)
dat$id <- as.factor(formatC(a, width = 2, flag = "0"))
dat$group <- as.factor(as.numeric(!is.nan(dat$diseaseAge)))
dat$sex <- as.factor(c("Male", "Female")[as.numeric(dat$sex)])
dat$group <- as.factor(c("Control", "Case")[as.numeric(dat$group)])
plot_sim(sim)
plot_sim(sim, comp_idx = 2)
```

## Visualizing longitudinal data
In this tutorial we use simulated `testdata_002`, which consists of 12 individuals and 8 time points for each. 

``````{r head}
data_info(testdata_002)
```

The `plot_data` function
can be used to visualize the data in many ways.

``````{r pd1, fig.width=7.2, fig.height=4.8}
plot_data(testdata_002, facet_by = "id", color_by = "sex") + xlab('Age (months)')
```

Coloring according to the disease-related age (`diseaseage`) shows that individuals 01-06 are cases and 07-12 are controls. The observed disease onset is at around 60
months for each case individual.
``````{r pd2, fig.width=7.2, fig.height=4.8}
plot_data(testdata_002, facet_by = "id", color_by = "diseaseAge") +
  scale_color_gradient2() + xlab('Age (months)')
```


## Creating and fitting a model
We define a model with four components and fit it. The `gp(age)` component
is a shared age effect, `zs(sex)*gp(age)` is a sex-specific deviation from it,
`gp_vm(diseaseAge)` is a non-stationary variance-masked disease effect, and
`zs(group)` is a static offset component between the two groups (Case/Control).

```{r fit, cache=TRUE}
fit <- lgp(formula = y ~ gp(age) + zs(sex)*gp(age) + gp_vm(diseaseAge) + zs(group),
           data     = testdata_002,
           iter     = 1000,
           chains   = 4,
           refresh  = 250)
```

## Studying the fit object
Printing the model object gives info about the model components, variables
and parameter priors. In this example we did not specify the `prior` argument when
calling `lgp()` so default priors were used.
```{r study1}
fit@model
```

Printing the fit object gives info about the posterior distribution of the
parameters, MCMC settings and convergence diagnostics.
```{r study2}
fit
```

Distribution of the parameter draws can be visualized. There are four magnitude
parameters (alpha), one for each component, and three lengthscale parameters
(ell), since the `zs(group)`component does not have a lengthscale parameter.
```{r study3, fig.width = 6, fig.height = 3.5}
plot_draws(fit, type = 'dens')
```

## Visualizing inferred components
```{r vis1, fig.width=3.5, fig.height=2}
plot_f(fit, comp_idx = 1, alpha_err = 0.1) + ylim(-1.8,1.8)
plot_f(fit, comp_idx = 2, color_by = "sex") + ylim(-1.8,1.8)
plot_f(fit, comp_idx = 3, color_by = "diseaseAge") + ylim(-1.8,1.8)
plot_f(fit, comp_idx = 4, color_by = "group") + ylim(-1.8,1.8)
```

```{r vis2, fig.width=5, fig.height=3}
plot_pred(fit)
```

## Out-of-sample predictions

