---
title: "Basic usage of the lgpr package"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
#  pdf_document:
#    toc: true     
vignette: >
  %\VignetteIndexEntry{Basic usage of the lgpr package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

This vignette demonstrates how to model a longitudinal data set and perform covariate selection using the `lgp` function of the `lgpr` package. In the example cases, the number of data points is small enough so that the experiments can be run in just a few seconds.

```{r load}
require("lgpr")
set.seed(123)
```

## Example
Here we simulate data with 8 individuals. We simulate individual-specific and shared age effects, as well as a sex-specific age effect. Also a categorial 'location' covariate is simulated, but it has no effect.

```{r example, fig.width=6.6, fig.height=4}
# generate artificial data -----------------------------------------------
simData <- simulate_data(N           = 8,
                        t_data       = seq(6, 30, by = 6),
                        covariates   = c(    2,2), # covariate types (2 = categorical)
                        lengthscales = c(6,6,6,6), # true effect lengthscales
                        relevances   = c(1,1,1,0), # true covariate relevances
                        names        = c("sex", "location"),
                        t_jitter     = 0.5,  # jitter in time points
                        snr          = 5)    # signal-to-noise ratio

data <- simData$data # data frame
plot_simdata(simData, componentwise = F)
plot_simdata(simData, componentwise = T)

# create and fit a model --------------------------------------------------
fit <- lgp(formula  = y ~ id + age + sex + location, # model formula
           data     = data,
           iter     = 1000, 
           chains   = 1,
           refresh  = 500)

# visualization -----------------------------------------------------------
plot(fit)
plot_components(fit)

# prediction --------------------------------------------------------------
#t_test <- seq(0, 40, length.out = 60)
#X_test <- create_test_points(fit, t_test)
#PRED   <- lgp_predict(fit, X_test, samples = "mean")
#plot_posterior_y(fit, PRED, uncertainty = "ribbon")
```

