---
title: "Basic usage of the lgpr package"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Basic usage of the lgpr package}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

This vignette demonstrates how to model a longitudinal data set and perform covariate selection using the `lgp` function of the `lgpr` package. In the example cases, the number of data points is small enough so that the experiments can be run in just a few seconds.

```{r load}
require("lgpr")
set.seed(123)
```

## Simulating data
Here we simulate data with 6 individuals. We simulate individual-specific and shared age effects, as well as a sex-specific age effect. Also a categorial 'location' covariate is simulated, but it has no effect.

```{r simulate}
sim <- simulate_data(N            = 6,
                     t_data       = seq(6, 30, by = 6),
                     covariates   = c(    2,2), # covariate types (2 = categorical)
                     lengthscales = c(6,6,6,6), # true effect lengthscales
                     relevances   = c(1,1,1,0), # true covariate relevances
                     names        = c("sex", "location"),
                     t_jitter     = 0.5,  # jitter in time points
                     snr          = 5)    # signal-to-noise ratio

sim
```

```{r plot_sim, fig.width=6.6, fig.height=4.7}
plot_sim(sim)
```

```{r components, fig.width=6.6, fig.height=5}
#plot(simdata, componentwise = TRUE)
```

## Creating and fitting a model
``````{r fit}
formula <- y ~ zs(id)*gp(age) + gp(age) + zs(sex)*gp(age) + zs(location)*gp(age)
fit <- lgp(formula  = formula,
           data     = sim@data,
           iter     = 1000, 
           chains   = 1,
           refresh  = 0)

show(fit@model)
```

## Visualizing inferred components and predictions
```{r vis, fig.width=6.6, fig.height=5}
#plot(fit)
#plot_posterior_components(fit)

# prediction
#t_test <- seq(0, 40, length.out = 60)
#X_test <- create_test_points(fit, t_test)
#PRED   <- lgp_predict(fit, X_test, samples = "mean")
#plot_posterior_y(fit, PRED, uncertainty = "ribbon")
```

