---
title: "Mathematical description of the statistical models in lgpr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mathematical description of the statistical models in lgpr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(lgpr)
```

This vignette describes mathematically the statistical models of `lgpr`, and 
shows which arguments of the `lgp()` modeling function customize which 
parts of the probabilistic model.

## Bayesian GP regression
The models in `lgpr` are models for the conditional distribution 
$$
p(y \mid f(\textbf{x}), \theta),
$$
of response variable $y$ given covariates $\textbf{x}$, where $\theta_{\text{obs}}$ is a possible parameter of the observation model
(like the magnitude of observation noise). The function $f$ has a GP prior 
$$
f \sim \mathcal{GP}(0, k\left(\textbf{x}, \textbf{x}' \mid \theta_{\text{GP}})\right),
$$

with covariance function $k(\textbf{x}, \textbf{x}' \mid \theta_{\text{GP}})$
that has hyperparameters $\theta_{\text{GP}}$. In addition to the GP prior for $f$, there is a parameter prior distribution $p(\theta)$ for $\theta = \left\{ \theta_{\text{GP}}, \theta_{\text{obs}} \right\}$. Given $N$ observations $\mathcal{D} = \{y_n, \textbf{x}_n\}_{n=1}^N$
the probabilistic models in `lgpr` have the form
\begin{align}
p\left(\theta, \textbf{f}\right) &= p\left(\textbf{f} \mid \theta\right) \cdot p(\theta) & \text{(prior)} \\
p(\textbf{y} \mid \textbf{f}, \theta) &= \prod_{n=1}^N p(y_n \mid f(\textbf{x}_n), \theta_{\text{obs}}) & \text{(likelihood)},
\end{align}
where $\textbf{f} = \left[ f(\textbf{x}_1), \ldots, f(\textbf{x}_N) \right]^{\top}$, $\textbf{y} = \left[y_1, \ldots, y_N\right]^{\top}$. The 
parameter prior density $p(\theta)$ is the product of the prior densities of each parameter, and the GP prior means that the prior for $\textbf{f}$ is the multivariate normal
\begin{equation}
p\left(\textbf{f} \mid \theta\right) = \mathcal{N}\left(\textbf{f} \mid \textbf{0}, \textbf{K} \right),
 \end{equation}
 where the $N \times N$ matrix $\textbf{K}$ has entries $\{ \textbf{K} \}_{in} = k(\textbf{x}_i, \textbf{x}_n \mid \theta_{\text{GP}})$.

## Model inference
After the model is defined, `lgpr` uses the MCMC methods of Stan to obtain draws
from the posterior
\begin{equation}
\label{eq: joint_posterior}
p\left(\theta, \textbf{f} \mid \mathcal{D}\right) \propto p\left(\theta, \textbf{f}\right) \cdot p(\textbf{y} \mid \textbf{f}, \theta) \\
\end{equation}

This requires evaluating the right-hand side of \ename~\ref{eq: joint_posterior} and its gradient thousands of times. 
 
If the observation model is Gaussian, $\textbf{f}$ can be analytically marginalized and only the marginal posterior $p\left(\theta \mid \mathcal{D}\right)$ needs to be sampled. This reduces the MCMC dimension by $N$ and likely improves sampling efficiency, but one however needs to evaluate $p(\textbf{y} \mid \theta)$ which is again an $N$-dimensional multivariate Gaussian.

## Additive GP regression

In additive GP regression, function $f = f^{(1)} + \ldots + f^{(J)}$ is modeled so that each component $j = 1, \ldots, J$ is a GP 
\begin{equation}
\label{eq: gp}
f^{(j)} \sim \mathcal{GP}\left(0, k_j(\textbf{x}, \textbf{x}') \right),
\end{equation}
independently from other components. This means that 
\begin{equation}
\label{eq: additive_gp}
f \sim \mathcal{GP}\left(0, k(\textbf{x}, \textbf{x}')  \right)  \hspace{0.7cm} \text{with} \hspace{0.7cm} k(\textbf{x}, \textbf{x}')  = \sum_{j=1}^J k_j(\textbf{x}, \textbf{x}’).
\end{equation}
and
\begin{equation}
\textbf{f}j{j} = \left[f^{(j)}(\textbf{x}_1), \ldots, f^{(j)}(\textbf{x}_N) \right]^\top  \sim \mathcal{N}\left(\textbf{0}, \textbf{K}j{j} \right)
\end{equation}
for each vector $\textbf{f}j{j}$, $j = 1, \ldots, J$. The matrix $\textbf{K}j{j}$ is defined so that its elements are $\{ \textbf{K}j{j} \}_{ik} = k_j(\textbf{x}_i,\textbf{x}_k)$. This means that the prior for $\textbf{f} = \textbf{f}j{1} + \ldots + \textbf{f}j{J}$ is $\textbf{f} \sim \mathcal{N}\left(\textbf{0},\textbf{K} \right)$, where $\textbf{K} = \sum_{j=1}^J \textbf{K}j{j}$.


