---
title: "Covariate uncertainty modeling in lgpr"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Covariate uncertainty modeling in lgpr}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

```{r load}
require("lgpr")
require("ggplot2")
```

## Simulating data

In this tutorial we generate a data set which contains a case and control
individuals, and the disease progression of case individuals is modeled using
the disease-related age as a covariate.
The true disease effect times for each case individual $q=1, \ldots,6$  are drawn from $\mathcal{N}(36,4^2)$, but the disease initiation is observable only after time $t_q$ , which is drawn from  $t_q∼\text{Exponential}(0.05)$ .

```{r simulate, fig.width = 7, fig.height=5}
set.seed(121)
relev           <- c(0,1,1,1,0,0)
effect_time_fun <- function(){rnorm(n = 1, mean = 36, sd = 4)}
obs_fun         <- function(t){min(t + stats::rexp(n = 1, rate = 0.05), 96 - 1e-5)}
  
simData <- simulate_data(N            = 12,
                         t_data       = seq(12, 96, by = 12),
                         covariates   = c(    0,2,2,2),
                         relevances   = relev,
                         lengthscales = c(18,24, 1, 18,18,18),
                         t_effect_range = effect_time_fun,
                         t_observed   = obs_fun,
                         snr          = 3)

plot_sim(simData)
```
Above, the blue line represents the data-generating signal and black dots are
noisy observations of the response variable.

```{r head}
dat <- simData@data
str(dat)
simData@effect_times
```

## Defining uncertainty prior

```{r prior}
lb <- 10
ub <- c(48,72,96,36,48,60)
unc_info <- list(zero = 0, backwards = FALSE, lower = lb, upper = ub)
```

```{r fit, fig.width=7.2, fig.height=4.8, cache = TRUE}
formula <- y ~ zs(id)*gp(age) + gp(age) + unc(id)*gp_vm(diseaseAge) + zs(z1)*gp(age) + zs(z2)*gp(age) + zs(z3)*gp(age)
fit <- lgp(formula   = formula,
            data     = dat,
            prior    = list(effect_time_info = unc_info),
            iter     = 20,
            chains   = 1,
            verbose  = TRUE)
```

```{r post, fig.width=7.2, fig.height=4.8}
print(fit)
print(fit@model)
```

