---
title: "Disease effect heterogeneity modeling in lgpr"
author: "Juho Timonen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: yeti
vignette: >
  %\VignetteIndexEntry{Disease effect heterogeneity modeling in lgpr}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#")
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

```{r load}
require("lgpr")
require("ggplot2")
require("rstan")
```

# Introduction

In this tutorial we simulate and analyse a test data set which contains 8 case and 8 control individuals, and the disease effect on case individuals is modeled using
the disease-related age (`diseaseAge`) as a covariate. The disease-related age is defined as age relative to the observed disease initiation. The disease effect is simulated so that
only 4 of the 8 case individuals are affected.

```{r simulate, fig.width = 7, fig.height=5}
set.seed(1213)
simData   <- simulate_data(N            = 16,
                           t_data       = seq(12, 72, by = 12),
                           covariates   = c(    0, 2,2,2),
                           relevances   = c(1,1,1, 1,0,0),
                           lengthscales = c(18,24, 1, 18,18,18),
                           t_effect_range  = c(46,48),
                           snr          = 5,
                           N_affected   = 4,
                           t_jitter     = 0)

plot_sim(simData)
#plot_sim.component(simData,3)
```
Above, the blue line represents the data-generating signal and black dots are
noisy observations of the response variable.

```{r head}
dat <- simData@data
str(dat)
simData@effect_times
```

# Declaring a heterogeneous component

We will define a formula where the term `het(id)*gp_vm(diseaseAge)`
declares that the `gp_vm` term is heterogeneous and one level-specific magnitude parameter is needed for each level of `id`.

```{r formula}
formula <- y ~ zs(id)*gp(age) + gp(age) + het(id)*gp_vm(diseaseAge) + zs(z1)*gp(age) + zs(z2)*gp(age) + zs(z3)*gp(age)
```

Because `diseaseAge` is `NaN` for the control individuals, it is automatically taken into account that an individual-specific magnitude parameter is actually needed just for each case individual.

# Fitting the model

```{r fit, fig.width=7.2, fig.height=5.4, cache = TRUE}
fit <- lgp(formula       = formula,
            data         = dat,
            prior        = list(wrp = igam(14,5)),
            iter         = 2000,
            chains       = 4,
            cores        = 4)
```

Printing the fit object summarizes the posterior
```{r post, fig.width=7.2, fig.height=4.8}
print(fit)
```

Printing the model information clarifies the model and priors
```{r mod}
model_summary(fit)
```

## Visualizing the individual-specific effect magnitudes
We can visualize the posterior distribution of the individual-specific disease effect magnitude parameters for each case individual. We see
that for individuals 1-4 the parameters are close to 1 (affected individuals)
and for individuals 5-8 there is considerable posterior mass close to 0.
```{r plt1, fig.width=7.2, fig.height=5.2}
plot_beta(fit)
```

# Visualizing the inferred heterogeneous effect
Finally we plot the inferred disease component, using posterior median
hyperparameters, and see that the inferred effects have the same shape for all individuals, but are heterogeneous in magnitude.

```{r plt2, fig.width=7.2, fig.height=4}
t <- seq(0, 100, by = 1)
x_pred <- new_x(dat, t, x_ns = "diseaseAge")
p <- pred(fit, x_pred, verbose = FALSE, reduce = stats::median)
plot_f(fit, pred = p, comp_idx = 3, color_by = 'diseaseAge')
```

# Computing environment

```{r sess}
sessionInfo()
```